Parallel Adaptive Importance Sampling

### 2016-04-08 PAIS.tex
Added ensemble size analysis and improved some numerics discussion

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the proposal distribution is over-dispersed, then the algorithm will propose states a long way from the current state at the cost of a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the space so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described
above, can be found by tuning the variance of the proposal
distribution $\nu$ during the burn-in phase of the
algorithm. Algorithms which use this method to find optimal proposal
distributions are known as adaptive MCMC algorithms. Adaptive MCMC
algorithms will converge to the stationary distribution $\pi(\cdot)$,
irrespective of whether the adaptive parameter itself converges to an
optimal value, under the following
conditions\cite{roberts2007coupling,roberts2009examples}: 
\begin{enumerate}
\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
\end{enumerate}

The variance of the proposal distributions that are used within the
proposal mixture distribution plays a key role in how well the
proposal distribution represents the target distribution, and
therefore how quickly the algorithm converges. A large over-dispersed
proposal distribution will sample a lot from the tails of the
distribution, but may also find other regions of high density with
respect to the target distribution. A very small variance in the
proposal distributions will lead to a very rough proposal mixture
distribution. In practise, we wish to find the happy medium in which
the proposal distribution is slightly more dispersed than the target
density, whilst also being a good representation of the regions which
have high density with respect to the target distribution.

Adaptively choosing the variance of the proposal distributions with a
large initial guess allows us to first explore the state space
globally, searching for multiple modes. Reducing the proposal variances to an
optimal value then allows us to explore each region efficiently. The
fact that we have multiple chains allows us to assess quickly and
effectively what the optimal variance of the proposal distributions
should be.

In many MCMC algorithms such as the Random Walk Metropolis-Hastings (RWMH) algorithm, the optimal scaling parameter can be found by searching for the parameter value which gives an optimal acceptance rate, e.g. for near Gaussian targets we have 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}. Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\delta$. Section~\ref{sec:statistics} gives some possible methods of tuning $\delta$.

@@

\subsubsection{Determining optimal scaling parameter $\delta$ using error analysis}

MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution. To assess this, a distance metric must be chosen. Examples are the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram. The relative error in the $m$-th moment is
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$ produced by the algorithm. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the points $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of a bin, and $B_i$ is the value of the $i$th bin.

These statistics are not practical for finding optimal values of
$\delta$ since they require knowledge of the analytic solution, and
require that the algorithm be run for a long time to build up a
sufficiently large sample. However they can be used to assess the
ability of other statistical measures to find the optimal proposal
variances in a controlled setting.

Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

The following statistics can be used for importance sampling algorithms.

\subsubsection{Determining optimal $\delta$ using the variance of the weights}

Importance samplers assign a weight to each sample they produce, based on a ratio of the posterior to the proposal at that point. This type of sampler is most efficient when the posterior is proportional to the proposal distribution, in this case the weights are constant, and so the variance of the weights, $var(w(y))$, is zero. Hence, the optimal value of $\delta$, is
\[
	\delta^*_{\text{var}} = \arg\min_{\delta} \text{var}(w(y)).
\]
The mean of the estimator $\text{var}(w(y))$ is a smooth function, and so it is used to tune $\delta$ during the burn-in phase of MCMC algorithms. The variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of samples to calculate descent directions.

\subsubsection{Determining optimal $\delta$ using the effective sample size}\label{sec:ess}

The effective sample size, $\neff$, can be used to assess the efficiency of importance samplers. Ideally, in each iteration, we would like all $M$ of our samples to provide us with new information about the posterior distribution. In practise, we are unlikely to achieve a ratio of exactly 1.

@@

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{figures/neff-M}
\caption{Left: A contour plot showing how $\delta_{\text{eff}}$ varies with the ensemble size. The black line highlights the optimal value of $\delta$. Right: The value of the effective sample size ratio for the optimal $\delta$ at each ensemble size. i.e. the value of the effective sample size ratio along the black line in the left figure. Set up for this problem is given in Section~\ref{sec:problem 2}.}
\label{fig:neff-M}
\end{center}
\end{figure}

If we are to use the effective sample size ratio as an indicator of
how to tune the proposal variance, we need to look at how it behaves
in different situations. In Section \ref{Sec:Num} we see how it behaves for different observation operators. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in \ref{sec:problem 2} when using the PAIS-MALA algorithm. The left part of the figure shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

@@

A popular approach for adaptive MCMC algorithms is to view the scaling
parameter as a random variable which we can sample during the course
of the MCMC iterations. However, it can be slow to converge to the
optimal value, and we may need an
uninformative prior for the scaling parameter. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain, a benefit being that it allows some exploration of the state space but never converges to an optimal value of the scaling parameter. We choose to use a divide and conquer scheme which optimises the effective sample size. Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. The proposed algorithm, is not presented as an optimal strategy but as an example of the benefits of tuning the algorithm using the effective sample size statistic.

From hereon in, we will assume that we are using the proposal from the
pCNL algorithm given in Section \ref{Sec:pCNL}, which is given by:
\begin{equation}Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
2\delta\mathcal{C}\nabla \Phi(u)+
\sqrt{8\delta} W\right], \qquad W \sim \mu_0,
\end{equation}
where $X$ is our current state and $Y$ is our proposed state. The
scaling parameter $\delta$ dictates the variance in the proposal distribution.

 In the adaptive algorithm described in Table~\ref{tab:adapt} in
 Appendix \ref{Sec:App}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations at which to update $\delta^{(j)}$, $\{n_k\}_{k=1}$, which can be decided using a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate $\delta^*$ for the pCNL algorithm.

When implemented, there are a number of parameters which must be
chosen to allow efficient tuning of $\delta$. Firstly, the initial
value of $\delta$ should be chosen so that the chains spread out
quickly across the state space. However, if it is chosen to be too
large, then the method may become unstable and inefficient. We also choose two iteration numbers at which we change our adaptive strategy. At iteration $N_\text{join}$ we resample using the full ensemble size instead of using to subsets of the full ensemble. This allows us to fine-tune the scaling parameter using the full resampler, but slows down the rate of convergence of the scaling parameter. Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter to guarantee ergodicity of the MCMC algorithm.

@@

In this example we look at how PAIS compares to naively parallelised
MH. We compare the RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We consider several statistics
for measuring the efficiency of the PAIS algorithm compared to
existing Metropolis-Hastings algorithms when applied to a simple one
dimensional Gaussian posterior.

Since we are comparing against naively parallelised MCMC algorithms,
we also need to decide which statistic $T(\delta)$ to optimise for
these approaches. In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors. To optimise the algorithm we minimise the statistic
\[
	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\delta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) =
\neff(\delta)$.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_\Sigma - \frac{1}{2}\big\|u\big\|^2_{\mathcal{T}}\right),
\end{equation}
where $\Sigma$ and $\mathcal{T}$ are the covariances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\Sigma = \sigma^2I_N$ and
$\mathcal{T} = \tau^2I_d$ with $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\Sigma).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_Y || \mu_0) = 4.67$ for this problem.

@@

{\bf (1) Finding the optimal parameters}: To find the optimal
parameters we choose 32 values of $\delta$ evenly spaced on a log scale
in he interval $[10^{-5}, 2]$. We run the PAIS-RW algorithm for 100,000 iterations and RWMH for 1,000,000 iterations, each with $M=50$ processes. We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for 1 million iterations, again with 50 processes. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative L2 error statistic in Equation~\ref{eqn:L2_error}.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative L2 error to compare
efficiency. The initial value of $\delta$ is given in the discussion
of each simulation.

@@

Both adaptive algorithms are run with initial values of $\delta_0=2.0$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth making it difficult to find the optimal parameter.

@@

Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales if we were to increase the ensemble size, and if there is some limit below which the algorithm fails. We implement the problem in Section~\ref{sec:problem 1}, using the pCNL algorithm with ensemble sizes ranging from $M=2$ up to $M=494$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-pCNL samples required to reach the same tolerance as the pCNL algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\delta$ described in Section~\ref{sec:Implementation P1}(1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naive
MCMC. When $M<8$, the algorithm takes a long time to reach
stationarity. This indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism.

@@

In this section we investigate the behaviour of the PAIS algorithm
when applied to bimodal problems. Metropolis-Hastings methods can
struggle with multimodal problems, particularly where switches between
the modes are rare, resulting in incorrectly proportioned modes in the
histograms, for example. With the PAIS algorithm, we see that the
resampling step redistributes chains to new modes as they are
found. This means that we expect the number of chains in a mode to be
approximately proportional to the probability mass in that mode. As a result, reconstructed posteriors with disproportional modes, as is familiar with the Metropolis-Hastings algorithms, are not produced. We again look at an `easy' problem, BM(1), which has a KL divergence of 0.880, and a `harder' problem, BM(2), which has a KL divergence of 3.647. Problem BM(1) has two modes which are separated by a smaller energy barrier. In BM(2) we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

@@

The following setup is the same for both problems. We consider an observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results in the non-Gaussian posterior
\[
	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
\]
To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$. In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point.

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problem BM(1) and BM(2). Problem BM(1) has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
\label{fig:problem 3 posteriors}
\end{center}repo
\end{figure}

@@

The numerical implementation for most of the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}. The only exception is that the convergence plots for both adaptive and nonadaptive algorithms are run for $10^6$ iterations instead of $10^7$.

@@

Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for pCNL and we use the effective sample size to find the optimal values for PAIS-pCNL. Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.

>>>>

Efficient selection of scaling parameters in MCMC algorithms is critical to achieving optimal mixing rates and hence achieving fast convergence to the target density. One significant difficulty is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the proposal distribution is widely over-dispersed, then the algorithm will propose states a long way from the current state resulting in a relatively low acceptance rate, impacting the quality of inferences on the posterior. Similarly, if the posterior is under-dispersed, the process will take a long time to fully explore the parameter space and so the mixing rate, and hence convergence rate, will be slow. It is therefore necessary to find a proposal distribution which is slightly over-dispersed to ensure the entire posterior is explored\cite{gelman1992inference}, but is as close to the posterior as possible.

Proposal distributions which are slightly over-dispersed as described
above, can be found by tuning the variance of the proposal
distribution $\nu$ during the burn-in phase of the
algorithm. Algorithms which use this method to find optimal proposal
distributions are known as adaptive MCMC algorithms. Adaptive MH
algorithms will converge to the stationary distribution $\pi(\cdot)$,
irrespective of whether the adaptive parameter itself converges to an
optimal value, under the following
conditions\cite{roberts2007coupling,roberts2009examples}: 
\begin{enumerate}
\item The adaptive parameter exhibits diminishing adaptation, which says that the amount of movement in the adaptive parameter decreases with the length of the chain. 
\item The smallest number of steps for which the kernel has sufficiently converged, from an initial state $x$, is finite, by bounded convergence.
\end{enumerate}

Algorithms which use mixture proposals, e.g. PAIS, must tune the variance of the individual kernels within the proposal mixture. Adaptively choosing the variance of the kernels starting with a
large initial guess allows us to first explore the state space
globally, searching for the locations of the modes. Reducing the proposal variances to an
optimal value then allows us to explore each region efficiently. The
fact that we have multiple chains allows us to assess quickly and
effectively what the optimal variance of the proposal distributions
should be. This is discussed in Section~\ref{sec:adapt}.

In many MCMC algorithms such as the Random Walk Metropolis-Hastings (RWMH) algorithm, the optimal scaling parameter can be found by searching for the parameter value which gives an optimal acceptance rate, e.g. for near Gaussian targets we have 23.4\% for RWMH and 57.4\% for MALA\cite{roberts2001optimal}. Unlike Metropolis-Hastings algorithms, the PAIS algorithm does not accept or reject proposed values, so we need another method of measuring the optimality of $\delta$. Section~\ref{sec:statistics} gives some possible methods for tuning $\delta$.

@@

\subsubsection{Determining optimal scaling parameter using error analysis}

MCMC algorithms can be assessed by comparing their approximation of the posterior to the analytic distribution. To assess this, a distance metric must be chosen. Examples are the relative error between the sample moments and the true moments, or the relative L2 error between the true density, $\pi(x|D)$, and the constructed histogram. The relative error in the $m$-th moment is
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative L2 error between a continuous function to a piecewise constant function, $e$, is
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional histogram bins, so that $\bigcup_i R_i \subseteq X$ and $R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the volume of each bin, and $B_i$ is the value of the $i$th bin.

These statistics cannot be used in practise to find optimal values of
$\delta$ since they require knowledge of the analytic solution, and
the algorithm must be run for a long time to build up a
sufficiently large sample. However they can be used to assess the
ability of other indicators to find the optimal proposal
variances in a controlled setting.

%%% remove this paragraph?
Another related statistic, is the variance of the sample estimate of the first moment $\hat{\mu}$. Assuming that the algorithm is converging to the invariant distribution, convergence will be fastest when the variance of the estimate is minimised. This fact follows from the convergence rate of Monte Carlo algorithms, $\sigma/\sqrt{n}$. The proposal distribution which minimises the variance, $\sigma^2$, of $\hat{\mu}$, is the most desirable. This variance statistic often converges faster than the moment itself, and does not require any knowledge of the true value of the moment.

The following statistics can be used specifically for importance sampling algorithms.

\subsubsection{The variance of the weights}

Importance samplers assign a weight to each sample they produce based on a ratio of the posterior to the proposal at that point. This method is most efficient when the target is proportional to the proposal distribution, in this case the weights are constant and so the variance of the weights, $var(w(y))$, is zero. Hence, we would like to choose the value of $\delta$ which minimises the variance of the weights,
\[
	\delta^*_{\text{var}} = \arg\min_{\delta} \text{var}(w(y)).
\]
The mean of the estimator $\text{var}(w(y))$ is smooth enough that it can be used to tune $\delta$ during the burn-in phase of MCMC algorithms. However the variance of the estimator can be large, especially far away from the optimal value, so it can take a large number of iterations to calculate descent directions.

\subsubsection{The effective sample size}\label{sec:ess}

The effective sample size, $\neff$, can also be used to assess the efficiency of importance samplers. Ideally, in each iteration, we would like all $M$ of our samples to provide us with new information about the posterior distribution. In practise, we cannot achieve a perfect effective sample size of $M$.

@@

\begin{figure}[htb]
\centering
\subfigure[Contours showing optimal ranges of the scaling parameter.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM_contour"}}
\subfigure[The highest effective sample size achieved for each ensemble size.]{\includegraphics[width=0.47\textwidth]{"figures/ESSvM"}}
\caption{The behaviour of the effective sample size as the ensemble size increases. This analysis is from the Gaussian posterior in Section~\ref{sec:problem 1} using the PAIS-RW algorithm.}
\label{fig:neff-M}
\end{figure}

If we are to use the effective sample size ratio as an indicator of
how to tune the proposal variance, we need to look at how it behaves
in different situations. Here we look at how the statistic behaves as we vary the ensemble size, $M$. Figure~\ref{fig:neff-M} shows results for the Gaussian posterior discussed in Section~\ref{sec:problem 1} when using the PAIS-RW algorithm. Figure~\ref{fig:neff-M} (a) shows that as the ensemble size increases, the value of $\delta$ which gives the optimal effective sample size ratio decreases. This is to be expected since we are trying to fit more kernels into the posterior distribution, so the variance of each kernel needs to be decreased.

Figure~\ref{fig:neff-M} (b) shows that as the ensemble size increases, the efficiency of the sampler also increases.

@@

A popular approach for adaptive MCMC algorithms is to view the scaling
parameter as a random variable which we can sample during the course
of the MCMC iterations. However, it can be slow to converge to the
optimal value, and we may need an
uninformative prior for the scaling parameter. Alternatively, the parameter may be randomly sampled from some interval at various points during the chain. This results in some iterations which explore the state space, and some which make local moves. Algorithms of this type do not converges to an optimal value of the scaling parameter.

We choose to use a divide and conquer scheme which optimises the effective sample size (or any other diagnostic). Some more sophisticated examples are described in \cite{roberts2009examples} and \cite{Ji2013adaptive}. Algorithm~\ref{tab:adapt} in Appedix~\ref{Sec:App}, is not presented as an optimal strategy but as a demonstration that the PAIS algorithm can be adaptively tuned effectively.

From here on in, we will use the random walk proposal,
\begin{equation}Y = X_{i-1}+\beta W_{i-1}, \qquad W_{i-1} \sim \mathcal{N}(0,1),
\end{equation}
where $X_{i-1}$ is our current state, $Y$ is our proposed state, and $\beta = \sqrt{8\delta}/(2+\delta)$. In the numerics section we will refer to $\delta$ as the scaling parameter.

 In Algorithm~\ref{tab:adapt}, we calculate a sequence $\{\delta^{(k)}\}_{k=1}$ which converges roughly to the optimal scaling parameter $\delta^*$, resulting in the optimal transition density $\chi$ for our MCMC algorithm. We must choose some sequence of iterations, $\{n_k\}_{k=1}$, at which to update $\delta$. These $n_k$ can be constructed by a sequence in which the terms grow exponentially further apart. The algorithm is given for the PAIS method but a similar method is used to adaptively calculate the optimal scaling parameter, $\delta^*$, for the MH algorithm.

When implemented, there are a number of parameters which must be
chosen to ensure efficient tuning of $\delta$. Firstly, the initial
value of $\delta$ should be chosen so that the chains quickly explore the parameter space. However, if $\delta$ is chosen to be too
large, then the method may become unstable. We also choose two iteration numbers at which we change our adaptive strategy. At iteration $N_\text{join}$ we resample using the full ensemble instead of using two subsets of the full ensemble. When using two subsets we are training the scaling parameter using the wrong ensemble size, this allows faster convergence but leads to a sub-optimal scaling parameter. Resampling with the full ensemble towards the end of the tuning allows fast convergence to the right order of magnitude, followed by fine tuning to the optimal scaling parameter. Finally we choose a time, $N_\text{stop}$, at which we stop updating our parameter.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. We perform this comparison using the relative L2 error defined in Equation~\ref{eqn:L2_error}, as well as the relative error in the first moment.

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistic $T(\delta)$ to optimise for
these approaches. In the examples which follow, we have optimised the naively parallelised RWMH algorithm using the optimal acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical asymptotic value of 0.234 which applies in higher dimensions, this higher acceptance rate is commonly used for one dimensional Gaussian posteriors. To optimise the algorithm we minimise the statistic
\[
	T_{\text{MH}}(\delta) = \left| \frac{N_{\text{acc}}(\delta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\delta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we are maximising the effective sample size as
discussed in Section~\ref{sec:ess}, so $T_{\text{PAIS}}(\delta) =
\neff(\delta)$.

@@

Consider the simple case of a linear observation operator $\G(u) = u$, where the prior on $u$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_Y) = \pi(u|D) \propto \exp\left(-\frac{1}{2}\big\|u - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|u\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $u=4$ noisily such that
\[
D = \mathcal{G}(u) + \eta \sim \mathcal{N}(\mathcal{G}(u),\sigma^2).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_Y || \mu_0) = 4.67$ for this problem.

@@

{\bf (1) Finding the optimal parameters}: To find the optimal
parameters we choose 32 values of $\delta$ evenly spaced on a log scale
in the interval $[10^{-5}, 2]$. We run the PAIS-RW and RWMH algorithms for one million iterations, each with an ensemble size of $M=50$. We took 32 repeats of both algorithms and then used the geometric means of the sample statistics to find the optimal parameters.

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative L2 error statistic in Equation~\ref{eqn:L2_error}.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative L2 error to compare
efficiency. The adaptive algorithms are initialised with $\delta^{(1)}$ = 2.

@@

Both adaptive algorithms are run with initial values of $\delta_0=2.0$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

Throughout this paper, we use an ensemble size $M=50$, but it is interesting to see how the PAIS algorithm scales when we increase the ensemble size, and if there is some limit below which the algorithm fails. We implement the problem in Section~\ref{sec:problem 1}, using the RWMH and PAIS-RW algorithms with ensemble sizes in the inteveral $M \in [1, 160]$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.7\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\delta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively parallelised MH. When $M<8$, the algorithm takes a long time to reach
stationarity. The decreaseing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

@@

In this section we investigate the behaviour of the PAIS algorithm
when applied to bimodal problems. MH methods can
struggle with multimodal problems, particularly where switches between
the modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the PAIS algorithm redistributes 
chains to new modes as they are found. This means that we expect the number of chains in a mode to be
approximately proportional to the probability density in that mode. As a result, reconstructed posteriors with disproportional modes, as is familiar with the MH algorithms, are not produced.

We look at an `easy' problem, $B_1$, which has a KL divergence of 0.880, and a `harder' problem, $B_2$, which has a KL divergence of 3.647. Problem $B_1$ has two modes which are separated by a small energy barrier. In $B_2$ we increase the distance between the two modes which has the effect of increasing the required energy to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

@@

The following setup is the same for both problems. We consider a nonlinear observation operator $\G(u) = u^2$, and assign the prior $u \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading, $D$, is taken according to $D = \G(u) + \varepsilon$, where $\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results in the non-Gaussian posterior
\[
	\pi(u|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|u^2 - D\|^2 - \frac{1}{2\tau^2}\|u\|^2\right).
\]
To create the `easy' problem we say that the true value of $\G(u) = 0.75$, and the `hard' problem is generated using $\G(u) = 2$. In the numerics which follow we draw noise from $\mu_\varepsilon$ to generate our data point.

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problem $B_1$ and $B_2$. Problem $B_1$ has a noisy data value of  0.921312 which results in an energy barrier which is relatively easy to cross, as well as their common prior distribution.}
\label{fig:problem 3 posteriors}
\end{center}
\end{figure}

@@

The numerical implementation for the following simulations follow the same setup as described in Section~\ref{sec:Implementation P1}.

@@

Calculating the optimal values of the scaling parameters for this problem is similar to the Gaussian case; we check only the acceptance rate to find the optimal values for RWMH and we use the effective sample size to find the optimal values for PAIS-RW. Table~\ref{table:BM_opt_delta} gives the optimal values of $\delta$ for both problems.

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$ \\ \hline
	RWMH								& 1.3e-1					& - \\
	PAIS-RW								& -						& 5.4e-3\\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\delta^*_{\text{acc}}$	& $\delta^*_{\text{eff}}$	& $\delta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.6e-2					& - 						& 9.1e-1\\
	PAIS-RW								& -						& 1.1e-3 					& 8.1e-3\\
	MALA								& 2.6e-2					& - 						& 1.3e-1\\
	PAIS-MALA							& -						& 3.7e-3 					& 8.1e-3\\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\delta$ for $B_1$ (left) and $B_2$ (right).}
	\label{table:BM_opt_delta}
\end{table}

Problem $B_2$ is much harder than $B_1$; transitions between the modes
are extremely unlikely for the standard pCNL algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find all the modes, and to sample them thoroughly and in the correct proportions.

To get correctly proportioned modes with the RWMH algorithm it is
important that the chains can transition between the modes, which
means that $\delta$ must be large. However, this leads to a lower
acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
converge for problems of this type.

We can achieve these two regimes in RWMH by tuning $\delta$ using the acceptance rate for local convergence, and by L2 error for global convergence. Similarly in PAIS-RW we can use the effective sample size for local convergence, and the L2 error for global convergence.
